---
title: Advanced R - Chapter 6 Notes
format:
  html:
      toc: true
---

# 6 Functions

## 6.2 Function Fundamentals

Section 6.2 describes the basics of creating a function, the three main components of a function, and the exception to many function rules: primitive functions (which are implemented in C, not R).

Functions are objects, just like vectors.

Parts of a function
- The `formals()`, the list of arguments that control how you call the function
- The `body()`, the code inside the function
- The `environment()`, the data structure that determines how the function finds the values associated with the names.

You specify the formals and body, but the environment is implicitly specified.

```{r}

f02 <- function(x, y) {
  # A comment
  x + y
}

formals(f02)
#> $x
#> 
#> 
#> $y

body(f02)
#> {
#>     x + y
#> }

environment(f02)
#> <environment: R_GlobalEnv>

```

Functions can posses attributes.

There is one exception to the rule that a function has three components. Primitive functions like `sum()` and `[` call C code directly.

```{r}

sum
#> function (..., na.rm = FALSE)  .Primitive("sum")
`[`
#> .Primitive("[")

# they are types 'builtin' or 'special'

typeof(sum)
#> [1] "builtin"
typeof(`[`)
#> [1] "special"

# because they exist in C, their formals, body, and environment are NULL

formals(sum)
#> NULL
body(sum)
#> NULL
environment(sum)
#> NULL

```


**First-class functions** just means R functions are objects. They don't require special syntax to define.

```{r}

f01 <- function(x) {
  sin(1 / x ^ 2)
}

```



Naming functions is not strictly necessary. Unnamed functions are called **anonymous functions**.

```{r}

lapply(mtcars, function(x) length(unique(x)))
Filter(function(x) !is.numeric(x), mtcars)
integrate(function(x) sin(x) ^ 2, 0, pi)

```

Functions can be contained in a list

```{r}

funs <- list(
  half = function(x) x / 2,
  double = function(x) x * 2
)

funs$double(10)
#> [1] 20

```

If you already have the arguments in a data structure, you can use `do.call()`

```{r}
args <- list(1:10, na.rm = TRUE)

do.call(mean, args)
#> [1] 5.5

```



## Function Composition

Section 6.3 discusses the strengths and weaknesses of the three forms of function composition commonly used in R code.

You can nest function calls or save intermediate results as variables. You can also pipe.

```{r}

square <- function(x) x^2
deviation <- function(x) x - mean(x)
x <- runif(100)

sqrt(mean(square(deviation(x))))
#> [1] 0.274

out <- deviation(x)
out <- square(out)
out <- mean(out)
out <- sqrt(out)
out
#> [1] 0.274


```

## Lexical scoping

Section 6.4 shows you how R finds the value associated with a given name, i.e. the rules of lexical scoping.

**Scoping** is finding the value associated with a name.

```{r}

x <- 10
g01 <- function() {
  x <- 20
  x
}

g01()

```

R uses **lexical scoping**: it looks up the values of names based on how a function is defined, not how it is called. "Lexical" indicates that the scoping rules use a parse-time, rather than a run-time structure.

R's lexical scoping follows four primary rules:
1. Name masking
2. Functions versus variables
3. A fresh start
4. Dynmamic lookup

Names defined inside a function mask names defined outside a function.

```{r}

x <- 10
y <- 20
g02 <- function() {
  x <- 1
  y <- 2
  c(x, y)
}
g02()
#> [1] 1 2

```


If a name isn't defined inside a function, R looks one level up.

```{r}

x <- 2
g03 <- function() {
  y <- 1
  c(x, y)
}
g03()
#> [1] 2 1

# And this doesn't change the previous value of y
y
#> [1] 20

```

The same rules apply for a function defined inside of another function.


```{r}

x <- 1
g04 <- function() {
  y <- 2
  i <- function() {
    z <- 3
    c(x, y, z)
  }
  i()
}
g04()

```


The same scoping rules apply to functions themselves, not just the arguments or body of a function.

```{r}

g07 <- function(x) x + 1
g08 <- function() {
  g07 <- function(x) x + 100
  g07(10)
}
g08()
#> [1] 110

```


A fresh start

```{r}

```

## Lazy evaluation

Section 6.5 is devoted to an important property of function arguments: they are only evaluated when used for the first time.

```{r}



```

## ... (dot-dot-dot)

Section 6.6 discusses the special ... argument, which allows you to pass on extra arguments to another function.

```{r}



```

## 6.7 Exiting a function

Section 6.7 discusses the two primary ways that a function can exit, and how to define an exit handler, code that is run on exit, regardless of what triggers it.

```{r}



```

## 6.8 Function forms

Section 6.8 shows you the various ways in which R disguises ordinary function calls, and how you can use the standard prefix form to better understand what’s going on.

```{r}



```

```{r}



```